---
title: 'Missing Data Estimation in Morphometrics: How Much is Too Much?'
author:
- Julien Clavel
- Gildas Merceron
- Gilles Escarguel
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
  fontsize: 8pt
  pdf_document:
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 2
---

***


This tutorial is the same as the one given in the Supplementary Material S2 in Clavel et al. (2014). We also provide a worked example on a simulated dataset.

Simplified tutorial for the use of Multiple Imputation (MI) techniques used in Clavel et al. (2014)
==

The following R functions requires the "missMDA", "mice", "Amelia", "NORM", and "Hmisc" packages (See Material & Methods - Clavel et al.(2014)).

```{r, eval=FALSE}
# Required packages
library(mice)
library(Amelia)
library(Hmisc)
library(missMDA)
library(norm)

# Load the dataset
data<-read.table("mydata.txt", sep="\t", dec=".", header=T)

# alternatively you can try the following function:
# data <- read.table(file.choose(), sep="\t", dec=".", header=T)

# Number of multiple imputations
m=20

# ---------------- Imputation of the dataset ---------------- #
#          just select one of these methods below             #
# ------------------------------------------------------------#

# Mice – Method = "pmm" (predictive mean matching) or "norm"
imp <-   mice(data,  m  =  m, method="pmm")

# MI-PCA – ncp = number of dimensions to use for the imputation process
imp <-  MIPCA(data,  ncp  =  2,  scale  =  TRUE,  method  =  "Regularized",  nboot  = m)

# Hmisc – (type= "pmm" or "regression") A,B,C,D,E are the data colnames
imp<-aregImpute(~A+B+C+D+E,data,n.impute=m,type="pmm",match="weighted")

# Amelia II
imp<-  amelia(data, m  =  m)

# Norm
data<-as.matrix(data)
preA <- prelim.norm(data)
datA <- em.norm(preA) #find the MLE for a starting value
rd<-trunc(1000000*runif(1) + 10)
rngseed(rd)
imp<-list()

  for (i in 1:m){
    impA <- da.norm(preA,datA,steps=50,showits=FALSE) # take 50 steps
    imp[[i]] <- data.frame(imp.norm(preA,impA,data))
  }

```


agglomerate.data & plot.MI functions
==

Based on m (>1) datasets imputed using a multiple imputation technique, the two following R functions average the m imputed datasets and display the 95% confidence ellipses associated to each specimen. The function to draw confidence ellipses is based on the R function “ELLI” proposed by Claude (2008). The following R functions requires the "missMDA", "mice", "Amelia", "NORM", "Hmisc" and "shapes" packages (See Material & Methods).

Combination of the results obtained with one of the MI methods (Mice, Amelia, Norm, MI-PCA, or Hmisc; see the "simplified tutorial above") is done using the "agglomerate.data" function. This function generates an averaged dataset (_agglomerate.data\$ImpM_), and a list with the m imputed datasets (_agglomerate.data\$Mi_).

**Example:**

```{r, eval=FALSE}
IM<-agglomerate.data(data=data, imp=imp, Mimp=20, Method="mice")
```


Where "data" is the dataset with the missing values, "imp" is the MI dataset object obtained with one of the MI methods (see above), "Mimp" is the number of MI, and "Method" is one of the methods described above ("mice", "norm", "hmisc", "missmda" or "amelia").


```{r, eval=FALSE}
plot.MI(IM, symmetric=TRUE, DIM=c(1,2), web=FALSE, ellipses=TRUE)
```

The "plot.MI" function allows the procrustes superimposition of the m imputed datasets onto the principal components calculated from the average MI-dataset. Symmetric = whether or not the matrices must be scaled to have unit sum of square. DIM = the dimensions to display on the biplot; web = whether or not the m imputed points for each specimen are linked to their related average MI-dataset points; ellipses = whether or not the 95% confidence ellipses around each specimen is drawn.

A worked example
==

```{r, message=FALSE, results='hide', echo=FALSE}
## Combination of the results

#Dependencies
library(shapes)
library(mice)
library(Amelia)
library(missMDA)
library(Hmisc)
library(norm)

# agglomerate.data (data,imp,Mimp,Method="mice")
#
# plot.MI(IM, symmetric=TRUE, DIM=c(1,2),web=FALSE,ellipses=TRUE)
#
#data= dataset with missing values
#
#imp=Imputed datasets
#
#Mimp=Number of MI
#
#Method= MI method used, must be "mice","amelia","hmisc","missmda" or “norm”
#
#Julien Clavel - 2013
#################################################
#################################################
##Agglomerate.data. Function to store the m imputed datasets and to compute the averaged dataset

agglomerate.data<-function(data,imp,Mimp,Method="mice"){
  
  Moy<-Mimp+1
  redata<-as.matrix(data)
  ximp<-array(redata,dim=c(nrow(redata),ncol(redata),Moy))
  #####################
  
  if(any(is.na(redata))==TRUE){
    if(Method=="mice" || Method=="amelia" || Method=="missmda" || Method=="hmisc" || Method=="norm"){
#####################MICE
      if(Method=="mice"){
  
        for(i in 1:Mimp){
          ximp[,,i]<-as.matrix(complete(imp,i))
          
        }
        ##Averaged dataset
        ximp[,,Moy]<-apply(ximp[,,1:Mimp],c(1,2),mean)
      }
      #####################
      #####################Amelia
      if(Method=="amelia"){
        for(i in 1:Mimp){
          ximp[,,i]<-as.matrix(imp$imputations[[i]])
          
        }
        ##Averaged dataset
        ximp[,,Moy]<-apply(ximp[,,1:Mimp],c(1,2),mean)
      }
      #####################
      #####################
      #####################NORM
      if(Method=="norm"){
        for(i in 1:Mimp){
          ximp[,,i]<-as.matrix(imp[[i]])
          
        }
        ##Averaged dataset
        ximp[,,Moy]<-apply(ximp[,,1:Mimp],c(1,2),mean)
      }
      #####################
      #####################MDA
      if(Method=="missmda"){
        
        for(i in 1:Mimp){
          ximp[,,i]<-as.matrix(imp$res.MI[,,i])
          
        }
        ##Averaged dataset
        ximp[,,Moy]<-apply(ximp[,,1:Mimp],c(1,2),mean)
      }
      ####################
      ####################Hmisc
      if(Method=="hmisc"){
        ##Extract the m data imputed for each variables
        ximp<-array(redata, dim=c(nrow(redata),ncol(redata),Moy))
        col<-1:ncol(redata)
        for(j in 1:ncol(redata)){
          if(sum(is.na(redata[,j]))==0){
            col<-col[-which(col==j)]
            next
          }
        }
        for(m in 1:Mimp){
          for(g in col){
            ximp[,,m][!complete.cases(ximp[,,m][,g]),g]<-imp$imputed[[g]][,m]
          }
        }
        ##Averaged dataset
        ximp[,,Moy]<-apply(ximp[,,1:Mimp],c(1,2),mean)
        
      }
    }else{
      ####################
      ##Warning messages
      cat("Error! You must indicate if you are using Mice, Amelia, missMDA, NORM, or Hmisc package","\n")
    }}else{
      ## Warning messages
      cat("There is no missing value in your dataset","\n")
    }
  #return(ximp)
  tabM<-ximp[,,Moy]
  colnames(tabM)<-colnames(redata)
  list("ImpM"=tabM,"Mi"=ximp[,,1:Mimp],"nbMI"=Mimp, "missing"=as.data.frame(redata))
}#End
###################################
########################################
###############################################
######################################################
##Function to draw confidence ellipses
ELLI<-function(x,y,conf=0.95,np)
{centroid<-apply(cbind(x,y),2,mean)
 ang <- seq(0,2*pi,length=np)
 z<-cbind(cos(ang),sin(ang))
 radiuscoef<-qnorm((1-conf)/2, lower.tail=F)
 vcvxy<-var(cbind(x,y))
 r<-cor(x,y)
 M1<-matrix(c(1,1,-1,1),2,2)
 M2<-matrix(c(var(x), var(y)),2,2)
 M3<-matrix(c(1+r, 1-r),2,2, byrow=T)
 ellpar<-M1*sqrt(M2*M3/2)
 t(centroid + radiuscoef * ellpar %*% t(z))}
############################### Function to Plot MI confidence ellipses using procruste superimposition
##################################
####################################
######################################

plot.MI<-function(IM,symmetric=FALSE,DIM=c(1,2),scale=FALSE,web=FALSE,ellipses=TRUE,...){
    if(any(is.na(IM$ImpM)==TRUE))
  { cat("There is still missing values in the imputed dataset, please check your imputation")
   break
  }else{
  Mo<-IM$nbMI+1
  pcaM<-princomp(IM$ImpM)
  cpdimM<-as.matrix(pcaM$scores[,DIM])   
  opa<-array(cpdimM,dim=c(nrow(cpdimM),ncol(cpdimM),Mo)) 
  for(i in 1:IM$nbMI){
    pca<-princomp(IM$Mi[,,i])
    opa[,,i]<-as.matrix(pca$scores[,DIM])
  }
  if(symmetric==TRUE){
    for (i in 1:IM$nbMI+1){
      trace<-sum(opa[,,i]^2)
      opa[,,i]<-opa[,,i]/sqrt(trace) 
    }
  }
  ############################ Ordinary Procrustes Analysis (library(shapes))
  for(k in 1:IM$nbMI){
    analyse<-procOPA(opa[,,Mo],opa[,,k], reflect=TRUE)
    opa[,,k]<-analyse$Bhat
  }
  opa[,,Mo]<-analyse$Ahat
  ######################## Principal component explained variance
  pvar<-pcaM$sdev^2
  tot<-sum(pvar)
  valX<-pvar[DIM[1]]
  valY<-pvar[DIM[2]]
  valX<-round(valX*100/tot,digits=2)
  valY<-round(valY*100/tot, digits=2)
  ######################## Plot function
  op <- par(no.readonly=TRUE)
  if(scale==TRUE){
    plot(opa[,1,Mo],opa[,2,Mo], type="p", pch=3, col=c(as.factor(ifelse(complete.cases(IM$missing) ==T, 1, 5))),lwd=1,xlim=range(opa[,1,Mo]),ylim=range(opa[,1,Mo]),xlab=paste("DIM",DIM[1],valX,"%",sep=" "),ylab=paste("DIM",DIM[2],valY,"%",sep=" "))
  }
  if(scale==FALSE){
    plot(opa[,1,Mo],opa[,2,Mo], type="p", pch=3, col=c(as.factor(ifelse(complete.cases(IM$missing) ==T, 1, 5))),lwd=1,xlab=paste("DIM",DIM[1],valX,"%",sep=" "),ylab=paste("DIM",DIM[2],valY,"%",sep=" "))
  }
  title("MI effect on Multivariate Analysis", font.main=3, adj=1)
  ## Store row names
  NR<-IM$missing
  rownames(IM$missing)<-NULL
  ##
  if(ellipses==TRUE){                                      
  coul<-as.numeric(rownames(IM$missing[complete.cases(IM$missing),]))
  for (j in coul){
  lines(ELLI(opa[j,1,],opa[j,2,],np=Mo), col="black", lwd=1)}
  coul<-as.numeric(rownames(IM$missing[!complete.cases(IM$missing),]))
  for (j in coul){
  lines(ELLI(opa[j,1,],opa[j,2,],np=Mo), col="red", lwd=1)}
  }else{ points(opa[,1,],opa[,2,],cex=0.5) }
  if(web==TRUE){
   coul<-as.numeric(rownames(IM$missing[complete.cases(IM$missing),]))
   for (j in coul){ 
   for(f in 1:IM$nbMI){
   segments(opa[j,1,Mo],opa[j,2,Mo], opa[j,1,f],opa[j,2,f], col="black", lwd=1) }
 } 
   coul<-as.numeric(rownames(IM$missing[!complete.cases(IM$missing),]))
  for (j in coul){ 
   for(f in 1:IM$nbMI){
   segments(opa[j,1,Mo],opa[j,2,Mo], opa[j,1,f],opa[j,2,f], col="red", lwd=1)}
 } 
 points(opa[,1,],opa[,2,],cex=0.5) 
 } 
 nom<-rownames(NR)
 text(opa[,1,Mo],opa[,2,Mo],nom, pos=1)
  
  abline(h=0,v=0, lty=3)
par(xpd=TRUE)  # Do not clip to the drawing area
lambda <- .025
legend(par("usr")[1], (1 + lambda) * par("usr")[4] - lambda * par("usr")[3],c("Complete", "Missing"), xjust = 0, yjust = 0,lwd=3, lty=1, col=c(par('fg'), 'red'))
par(op)      
}
}

```

We use a simulated example to illustrate how the method works:

```{r}
library(MASS)
# Just make a random dataset
set.seed(100)

# Simulate correlated traits
p=5; n=50
# covariance matrix
sigma <- t(matrix(value <- rnorm(p*p),p,p))%*%matrix(value,p,p) 
data <- mvrnorm(n, mu=rep(0,p), Sigma=sigma)
colnames(data) <- c("A","B","C","D","E")

# Introduce some, say 5, "NA" values
data[sample(1:(p*n), 5)] <- NA
```

Now we try to impute the missing cases using the _mice_ function with the "pmm" model (predictive mean matching).

```{r}
# First source the codes provided in the Supplementary Material S2 
# source("agglomerate_and_plot.r")

library(mice)
m = 50 # number of multiple imputations
imp <- mice(data,  m  =  m, method="pmm", print=FALSE)

# Combine the results
IM <- agglomerate.data(data=data, imp=imp, Mimp=m, Method="mice")

# Compare the dataset with NA to the averaged imputed dataset:
head(data)

head(IM$ImpM)

# Plot the MI datasets axis
plot.MI(IM, symmetric=TRUE, DIM=c(1,2), web=FALSE, ellipses=TRUE)
```



References
==

**Claude J. 2008**. Morphometrics with R. New York: Springer 

**Clavel J., Merceron G., Escarguel G. 2014**. Missing Data Estimation in Morphometrics: How Much is Too Much? Syst. Biol. 63:203-218.


